---
title: "Setup"
engine: knitr
---

## Material

{{< downloadthis ../assets/pdf/01_introduction.pdf dname="01_introduction" label="Download the presentation" icon="filetype-pdf" >}}

## Exercises

### Login and set up

Choose one of the following:

- **Enrolled:** if you are enrolled in a course with a teacher
- **Own installation:** if you want to install packages on your own local Rstudio installation
- **Docker:** if you want to use the docker image locally

::: {.panel-tabset}

## Enrolled

Log in to Rstudio server with the provided link, username and password.

## Own installation

Install the required packages using [renv](https://rstudio.github.io/renv/index.html)

```{r}
#| eval: false
install.packages("renv")

renv::restore(lockfile = "https://raw.githubusercontent.com/sib-swiss/single-cell-r-training/refs/heads/master/renv.lock")
```

## Docker

With docker, you can use exactly the same environment as we use in the enrolled course, but than running locally.

In the video below there's a short tutorial on how to set up a docker container for this course. Note that you will need administrator rights, and that if you are using Windows, you need the latest version of Windows 10.

{{< video https://vimeo.com/481620477 width="640" height="360" >}}

The command to run the environment required for this course looks like this (in a terminal):

::: {.callout-warning}
## Modify the script
The home directory within the container is mounted to your current directory (`$PWD`), if you want to change this behaviour, modify the path after `-v` to the working directory on your computer before running it.
:::

```{bash}
#| eval: false
docker run \
--rm \
-p 8787:8787 \
-e PASSWORD=test \
-v $PWD:/home/rstudio \
pfeiferl/training-singlecell-rstudio:latest
```

If this command has run successfully, approach Rstudio server like this:

```
http://localhost:8787
```

Copy this URL into your browser. If you used the snippet above, the credentials will be:

- **Username:** `rstudio`
- **Password:** `test`

Great! Now you will be able to use Rstudio with all required installations.

::: {.callout-note}
## About the options
The option `-v` mounts a local directory in your computer to the directory `/home/rstudio` in the docker container ('rstudio' is the default user for Rstudio containers). In that way, you have files available both in the container and on your computer. Use this directory on your computer. Change the first path to a path on your computer that you want to use as a working directory.

The part `pfeiferl/training-singlecell-rstudio:latest` is the image we are going to load into the container. The image contains all the information about software and dependencies needed for this course. When you run this command for the first time it will download the image. Once it's on your computer, it will start immediately.
:::


:::


### The course data and environment

First, source (run) provided `set_libpaths.R` script to point R session to specific location of R packages (libraries). This command has to be run each time an R session is created. Second command copies input data from shared location to current working directory. Data are retained over session restarts. These commands apply only to specific Open OnDemand environment.

```{r}
#| eval: false

source('/gpfs/helios/projects/elixir_scRNA-seq/set_libpaths.R')

system('cp -rv /gpfs/helios/projects/elixir_scRNA-seq/input_data/course_data/ .')
```

Have a look at the `course_data` directory. It should contain the following:
Mind you, in some cases the course_data directory might contain fastq reads too. 
```
course_data/
├── count_matrices
│   ├── ETV6-RUNX1_1
│   │   └── outs
│   │       ├── filtered_feature_bc_matrix
│   │       │   ├── barcodes.tsv.gz
│   │       │   ├── features.tsv.gz
│   │       │   └── matrix.mtx.gz
│   │       └── web_summary.html
│   ├── ETV6-RUNX1_2
│   │   └── outs
│   │       └── filtered_feature_bc_matrix
│   │           ├── barcodes.tsv.gz
│   │           ├── features.tsv.gz
│   │           └── matrix.mtx.gz
│   ├── ETV6-RUNX1_3
│   │   └── outs
│   │       └── filtered_feature_bc_matrix
│   │           ├── barcodes.tsv.gz
│   │           ├── features.tsv.gz
│   │           └── matrix.mtx.gz
│   ├── PBMMC_1
│   │   └── outs
│   │       └── filtered_feature_bc_matrix
│   │           ├── barcodes.tsv.gz
│   │           ├── features.tsv.gz
│   │           └── matrix.mtx.gz
│   ├── PBMMC_2
│   │   └── outs
│   │       └── filtered_feature_bc_matrix
│   │           ├── barcodes.tsv.gz
│   │           ├── features.tsv.gz
│   │           └── matrix.mtx.gz
│   └── PBMMC_3
│       └── outs
│           └── filtered_feature_bc_matrix
│               ├── barcodes.tsv.gz
│               ├── features.tsv.gz
│               └── matrix.mtx.gz
├── proB.rds
└── sample_info_course.csv

20 directories, 21 files

```

This data comes from:

Caron M, St-Onge P, Sontag T, Wang YC, Richer C, Ragoussis I, et al. **Single-cell analysis of childhood leukemia reveals a link between developmental states and ribosomal protein expression as a source of intra-individual heterogeneity.** Scientific Reports. 2020;10:1–12. Available from: http://dx.doi.org/10.1038/s41598-020-64929-x

We will use the reads to showcase the use of `cellranger count`. The directory contains only reads from chromosome 21 and 22 of one sample (`ETV6-RUNX1_1`). The count matrices are output of `cellranger count`, and we will use those for the other exercises in `R`.
