---
title: "Normalization and scaling"
editor_options: 
  chunk_output_type: console
---

::: {.callout-note}
## Learning outcomes

**After having completed this chapter you will be able to:**

- Describe and perform standard procedures for normalization and scaling with the package `Seurat`
- Select the most variable genes from a `Seurat` object for downstream analyses
::: 


## Material

## Material

{{< downloadthis ../assets/pdf/03_normalization_scaling.pdf dname="Dimensionality_Reduction_SIB" label="Download the presentation" icon="filetype-pdf" >}}

{{< video https://youtu.be/QS6ZH_41b4s?si=fvaNGkqjsIcxTEFn >}}

{{< downloadthis ../assets/pdf/scRNAseq_RM_Integration_dimreduction.pdf dname="scRNAseq_RM_Integration_dimreduction" label="Download the presentation from the video" icon="filetype-pdf" >}}

- [Seurat vignette](https://satijalab.org/seurat/articles/pbmc3k_tutorial.html)

## Exercises

```{r}
#| warning: false
#| message: false
library(Seurat)
library(glmGamPoi)
seu <- readRDS("day1/seu_day1-2.rds")
```


### Normalization

After removing unwanted cells from the dataset, the next step is to normalize the data.
By default, Seurat employs a global-scaling normalization method `"LogNormalize"` that normalizes the feature expression measurements for each cell by the total expression, multiplies this by a scale factor (10,000 by default), and log-transforms the result. This process is crucial to account for variations in sequencing depth across different cells, ensuring that differences in gene expression are biological rather than technical. The log-transformation helps to stabilize variance and makes the data more amenable to downstream statistical analyses, as it reduces the impact of highly expressed genes.
Normalized values are stored in the "RNA" assay (as item of the `@assay` slot) of the seu object.

This is how you can call the function (don't run it yet! Read the exercise first):

```{r}
# Don't run it yet! Read the exercise first
seu <- Seurat::NormalizeData(seu,
                     normalization.method = "LogNormalize",
                     scale.factor = 10000)
```


::: {.callout-important}
## Exercise
Have a look at the assay data before and after running `NormalizeData()`. Did it change?
::: 

::: {.callout-tip}
You can extract assay data with the function `Seurat::GetAssayData`. By default, the slot `data` is used (inside the slot `assay`), containing normalized counts. Use `Seurat::GetAssayData(seu, layer = "counts")` to get the raw counts. 
:::

::: {.callout-tip collapse="true"}
# Answer
You can check out some assay data with:

```{r}
#| eval: false
Seurat::GetAssayData(seu)[1:10,1:10]  
```

Returning:
    
::: {.panel-tabset}
## Before normalization

```{r}
#| echo: false
Seurat::GetAssayData(object = seu, layer = "counts")[1:10,1:10]  
```

## After normalization

```{r}
#| echo: false
seu <- Seurat::NormalizeData(seu,
                     normalization.method = "LogNormalize",
                     scale.factor = 10000)
Seurat::GetAssayData(object = seu, layer = "data")[1:10,1:10]  
```

:::
::: 

::: {.callout-note}
## Updating `seu`
As you might have noticed, this function takes the object `seu` as input, and it returns it to an object named `seu`. We can do this because the output of such calculations are added to the object, without loosing information.
:::

### Variable features

We next calculate a subset of features that exhibit high cell-to-cell variation in the dataset (i.e, they are highly expressed in some cells, and lowly expressed in others). Focusing on these genes in downstream analysis helps to highlight biological signal in single-cell datasets. The procedure in Seurat models the mean-variance relationship inherent in single-cell data, where highly expressed genes tend to have higher variance. By identifying genes with variance that is higher than expected given their mean expression, we can select features that are more likely to represent true biological variability rather than technical noise. This selection of variable features is crucial for dimensionality reduction techniques like PCA, as it focuses the analysis on the most informative genes.

```{r}
seu <- Seurat::FindVariableFeatures(seu,
                            selection.method = "vst",
                            nfeatures = 2000)
```

Let's have a look at the 10 most variable genes:

```{r}
# Identify the 10 most highly variable genes
top10 <- head(Seurat::VariableFeatures(seu), 10)
top10
```

We can plot them in a nicely labeled scatterplot:

```{r}
vf_plot <- Seurat::VariableFeaturePlot(seu)
Seurat::LabelPoints(plot = vf_plot,
            points = top10, repel = TRUE)
```

::: {.callout-warning}
## Make sure the plotting window is large enough

The function `LabelPoints` will throw an error if the plotting window is to small. If you get an error, increase plotting window size in RStudio and try again. 
::: 

You can see that most of the highly variables are antibody subunits (starting with IGH, IGL). Not very surprising since we look at bone marrow tissue. We can have a look later in which cells they are expressed. 

### Scaling

Next, we apply scaling, a linear transformation that is a standard pre-processing
step prior to dimensional reduction techniques like PCA. The `ScaleData()` function

1. shifts the expression of each gene, so that the mean expression across cells is 0
2. scales the expression of each gene, so that the variance across cells is 1

This step is crucial because without it, highly expressed genes would disproportionately influence the principal components, masking the contributions of less abundant but potentially biologically significant genes. By scaling, we ensure that all genes contribute more equally to the downstream dimensionality reduction, allowing for a more accurate representation of biological variation. The results of this are stored in `seu$RNA@scale.data`

```{r}
seu <- Seurat::ScaleData(seu)
```


::: {.callout-note}
## The use of `Seurat::SCTransform`

The functions `NormalizeData`, `VariableFeatures` and `ScaleData` can be replaced by the function `SCTransform`. The latter uses a more sophisticated way to perform the normalization and scaling, and is [argued to perform better](https://genomebiology.biomedcentral.com/articles/10.1186/s13059-019-1874-1). However, it is slower, and a bit less transparent compared to using the three separate functions. Therefore, we chose not to use `SCTransform` for the exercises.

:::


::: {.callout-important}
## Bonus exercise
Run `SCTransform` on the `seu` object. Where is the output stored?

::: 

::: {.callout-tip collapse="true"}
## Answer
You can run it like so:

```{r}
#| output: false
seu <- Seurat::SCTransform(seu)
```

And it will add an extra assay to the object. `names(seu@assays)` returns:

```{r}
#| echo: false
names(seu@assays)
```

Meaning that a whole new assay was added (including the sparse matrices with counts, normalized data and scaled data). 
:::

::: {.callout-warning}
Running `SCTransform` will change `@active.assay` into `SCT`(in stead of `RNA`; check it with `DefaultAssay(seu)`). This assay is used as a default for following function calls. To change the active assay to `RNA` run:

```{r}
DefaultAssay(seu) <- "RNA"
```
:::

### Save the dataset and clear environment

Now, save the dataset so you can use it tomorrow:

```{r}
saveRDS(seu, "day2/seu_day2-1.rds")
```

Clear your environment:

```{r}
#| eval: false
rm(list = ls())
gc()
.rs.restartR()
```

